---
title: "Course Project: Practical Machine Learning"
author: "Juan Pablo Loaiza Ramírez"
date: "12/1/2020"
output: html_document
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = FALSE, warning = FALSE, message = FALSE)
```

### Executive Summary
One thing that people regularly do is quantify how *much* of a particular activity they do, but they rarely quantify how *well they do it*. The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants, in order to predict the manner in which they did the exercise. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har.

### Loading libraries and data
First, the two datasets (training and testing) are loaded, together with the libraries that will be used.
```{r loadingData}
library(caret)
library(randomForest)
library(rattle)
library(rpart)
library(rpart.plot)
## Getting training and testing URL links
trainingLink <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testingLink <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
## Downloading data into the directory folder
download.file(trainingLink, destfile = "training.csv")
download.file(testingLink, destfile="testing.csv")
## Loading data into the workspace
training <- read.csv("training.csv", na.strings = c("NA", "#DIV/0!", ""))
testing <- read.csv("testing.csv", na.strings = c("NA", "#DIV/0!", ""))
```

### Cleaning data
Second, the datasets are cleaned from unnecessary information.
```{r cleaningData}
## Deleting NA predictors
training <- training[, colSums(is.na(training)) == 0]
testing <- testing[, colSums(is.na(testing)) == 0]
## Deleting unnecesary columns
training <- training[, -c(1:7)]
testing <- testing[, -c(1:7)]
```

### Partitioning training data
Third, the training dataset is partitioned into training and validation sets, setting a seed before to make the analysis reproducible.
```{r splitTrainingData}
set.seed(31721)
inTrain <- createDataPartition(training$classe, p = 0.7, list = F)
trainTrain <- training[inTrain, ]
trainValid <- training[-inTrain, ]
dim(trainTrain)
dim(trainValid)
```

### Modelling and cross validation
Before testing a final model, different machine learning models will be trained to see their accuracy.
```{r modelling}
## Modelling with regression trees
rpartModel <- train(classe ~ ., data = trainTrain, method = "rpart")
rpartPred <- predict(rpartModel, trainValid)
confusionMatrix(factor(trainValid$classe), rpartPred)
fancyRpartPlot(rpartModel$finalModel)
## Modelling with Linear Discriminant Analysis
ldaModel <- train(classe ~ ., data = trainTrain, method = "lda")
ldaPred <- predict(ldaModel, trainValid)
confusionMatrix(factor(trainValid$classe), ldaPred)
## Modelling with random forest
rfModel <- train(classe ~ ., data = trainTrain, method = "rf")
rfPred <- predict(rfModel, trainValid)
confusionMatrix(factor(trainValid$classe), rfPred)
```

### Testing the chosen model
According to the cross validation, the model with highest accuracy is random forest, although it takes a very long time to process. For that reason, the random forest model is used to predict on the test set.
```{r finalTest}
finalPred <- predict(rfModel, testing)
finalPred
```